---
title: "Final Project for Practical Machine Learning "
author: "Chris"
date: "April 11, 2017"
output: html_document
---
##Introduction
####Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

####Data comes from  http://groupware.les.inf.puc-rio.br/har

##Source of Data
####The training data for this project are available here:

####https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

####The test data are available here:

####https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

####The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.


```{r echo=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

##Load relevant Libraries
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
```

###Downloading Data set
```{r pressure, echo=TRUE, message=FALSE, warning=FALSE}
set.seed(12345)
trainingset <-"http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testingset<-"http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
train <- read.csv(url(trainingset),na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testingset),na.strings=c("NA","#DIV/0!",""))
````
###Susbet training and test sets and assessing structure 
```{r,echo=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
Train <- createDataPartition(train$classe, p=0.6, list=FALSE)
myTraining <- train[Train, ]
myTesting <- train[-Train, ]
dim(myTraining); dim(myTesting)
````
###Cleaning up structure of traing set removing any non zero variance variables that are useless predictors. 
````{r, echo=TRUE,message=FALSE, warning=FALSE}
n <- nearZeroVar(myTraining, saveMetrics=TRUE)
myTraining <- myTraining[,n$nzv==FALSE]

n<- nearZeroVar(myTesting,saveMetrics=TRUE)
myTesting <- myTesting[,n$nzv==FALSE]

myTraining <- myTraining[c(-1)]
````
#####Eliminate training set data that has data that has more than 60% NA in it. 
```{r,echo=TRUE,message=FALSE, warning=FALSE}
trainingV3 <- myTraining
for(i in 1:length(myTraining)) {
    if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .7) {
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) == 1)  {
                trainingV3 <- trainingV3[ , -j]
            }   
        } 
    }
}

myTraining <- trainingV3
rm(trainingV3)
```

````{r,echo=TRUE,message=FALSE, warning=FALSE}
clean1 <- colnames(myTraining) ##Make sure predictors in training and testing are the same 
clean2 <- colnames(myTraining[,-58])###Isolate classe column  
myTesting <- myTesting[clean1]###Enusre only columns in data are the same as the training 
testing <- testing[clean2]             
dim(myTesting)
dim(testing)
```
####Make sure class in testing set is the same as that in thre training set 
```{r,echo=TRUE,message=FALSE, warning=FALSE}
for (i in 1:length(testing) ) {
    for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) == 1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}
testing <- rbind(myTraining[2, -58] , testing)
testing <- testing[-1,]

```
###Prediction RUles using Decision Trees
```{r,echo=TRUE,message=FALSE, warning=FALSE }
set.seed(12345)
modA <- rpart(classe ~ ., data=myTraining, method="class")
table(myTraining$classe)
fancyRpartPlot(modA)

```
###Confusion matrix of Random Forest analysis 
````{r,echo=TRUE,message=FALSE, warning=FALSE}

A1 <- predict(modA, myTesting, type = "class")
tree <- confusionMatrix(A1, myTesting$classe)
tree
````
###Plot table of random forest predictions using the classe as a reference. 

```{r,echo=TRUE,message=FALSE, warning=FALSE }
plot(tree$table, col = tree$byClass, main = paste("Analysis by Decision Tree: Accuracy =", round(tree$overall['Accuracy'], 4)))
````
###Prediction of Data using Random Forest 
````{r,echo=TRUE,message=FALSE, warning=FALSE}
set.seed(12345)
modFitB1 <- randomForest(classe ~ ., data=myTraining)
predB1 <- predict(modFitB1, myTesting, type = "class")
cmf <- confusionMatrix(predB1, myTesting$classe)
cmf
````
###Plot of of Random FOrest analysis of class vs other predictors 
```{r,echo=TRUE,message=FALSE, warning=FALSE }
plot(modFitB1)
````
###Plot of Confusion matrix using random forest 
```{r, echo=TRUE,message=FALSE, warning=FALSE}
plot(cmf$table, col = cmf$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmf$overall['Accuracy'], 4)))
````
###The results show that the random forest prediction method is more accurate than the the decision tree method and so the predicition analysis for the test samples will be assessed using the Random forest method.

###Prediction Using Boosting Method 
````{r,echo=TRUE, message=FALSE, warning=FALSE}
set.seed(12345)
fitControl <- trainControl(method = "repeatedcv",
                           number = 5,
                           repeats = 1)

gbmFit1 <- train(classe ~ ., data=myTraining, method = "gbm",
                 trControl = fitControl,
                 verbose = FALSE)


gbmFinMod1 <- gbmFit1$finalModel

gbmPredTest <- predict(gbmFit1, newdata=myTesting)
gbmAccuracyTest <- confusionMatrix(gbmPredTest, myTesting$classe)
gbmAccuracyTest
````
###Plot comparing accuracy of boosting method vs. the boosting iterations. The greater iterations had more accuracy. 

```{r,echo=TRUE,message=FALSE, warning=FALSE}
plot(gbmFit1, ylim=c(0.9, 1))
````

###Since Accuracy of test samples is the highest with the Random Forest method, this will be used to predict the classe output of the test sample. 
````{r,echo=TRUE, message=FALSE, warning=FALSE}
B2 <- predict(modFitB1, testing, type = "class")
B2
````
####For loop is generated to print the classe output intop txt files. 
````{r,echo=TRUE,message=FALSE, warning=FALSE}
pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}

pml_write_files(B2)
````
##Conclusion 
####As shown by the analysis, the confusion matrix has shown that the random forest provides the best prediction model to assess the classe data from the data provided. 